Nextstrain: real-time tracking of pathogen evolution

### 2018-04-17 replyToReviewer.tex

\begin{quote}
3. Page 1, Line 42: it might be helpful to point the reader to a general reference on phylodynamics, assuming the reference limit hasn't already been reached.
\end{quote}
We have reworded this section to include a reference to ``Viral Phylodynamics'', Voltz \textit{et al.} 2013.

\begin{quote}
4. Page 1, Line 43 (column 2): could references to the three databases (papers or URLs) be included here?
\end{quote}
We have added URLs to the text.

\begin{quote}
5. Page 2, Line 35: if it can be explained in a very short sentence, it might be interesting to mention the caveat of "nearly all outbreaks" - which ones aren't amenable to Nextstrain-ification?
\end{quote}
There are indeed pathogens for which the general approach outlined in this paper is insufficient, and so we have changed the text to read ``Nextstrain is able to be extended to most outbreaks with readily accessible genomic data, although we note the potential for recombination or slow mutation rate to confound phylogenetic signal.''

\begin{quote}
6. Page 2, Line 51: it should be noted that the transmission inference depends on how well the outbreak in question was sampled - if data is missing, important transmission links could be missed.
\end{quote}
This is a very important point, and we have added the following sentence to the manuscript: ``It is important to note that sampling bias can obscure transmission links, and in certain cases we have chosen not to display the inferred states due to this bias.''

\begin{quote}
7. Page 2, Line 20 (column 20): is there an example of one of these mutations that could be mentioned here?
\end{quote}
We have added a reference to the 2017 publication ``Three mutations switch H7N9 influenza to human-type receptor specificit'' (de Vries \textit{et al.}), however have not listed these mutations as we detect none in the (consensus) sequences shown on nextstrain.org

### 2018-04-17 whitepaper.tex
point 6

Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probable transmission events given the sampled data, together with a marginal likelihood for each event.
Inferred transmissions are displayed as colours on the tree and each transmission is also drawn on the map.
It is important to make known the confidence of any such reconstruction, and we do this in two ways: by matching colour saturation to the confidence of that trait, and by displaying all relevant information when one hovers over the corresponding branch or isolate on the tree.

@@

For many pathogens, the emergence and spread of gain-of-function variants is a grave concern.
For instance, China has experienced seasonal epidemics of influenza A/H7N9 over the past five years, however human to human transmission is thought to be rare, with the majority of human cases being spillover from poultry.
The threat of mutations which facilitate human to human transmission is of extreme concern, as H7N9 has a mortality rate of around $30\%$ \citep{li2014epidemiology}.
% \textit{In vitro} experiments have identified three amino acid mutations in the hemagglutinin (HA) gene which confer a switch in specificity for human-type receptors by promoting binding to human trachea epithelial cells \citep{devries2017three}.
Nextstrain allows monitoring of experimentally determined mutations, such that interventions and control measures may be implemented where appropriate.

>>>>

Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probable transmission events given the sampled data, together with a marginal likelihood for each event.
It is important to note that sampling bias can obscure transmission links, and in certain cases we have chosen not to display the inferred states due to this bias.
Inferred transmissions are displayed as colours on the tree and each transmission is also drawn on the map.
It is important to make known the confidence of any such reconstruction, and we do this in two ways: by matching colour saturation to the confidence of that trait, and by displaying all relevant information when one hovers over the corresponding branch or isolate on the tree.

@@

For many pathogens, the emergence and spread of gain-of-function variants is a grave concern.
For instance, China has experienced seasonal epidemics of influenza A/H7N9 over the past five years, however human to human transmission is thought to be rare, with the majority of human cases being spillover from poultry.
The threat of mutations which facilitate human to human transmission is of extreme concern (e.g. see \citep{deVries2017}), as H7N9 has a mortality rate of around $30\%$ \citep{li2014epidemiology}.
% \textit{In vitro} experiments have identified three amino acid mutations in the hemagglutinin (HA) gene which confer a switch in specificity for human-type receptors by promoting binding to human trachea epithelial cells \citep{devries2017three}.
Nextstrain allows monitoring of experimentally determined mutations, such that interventions and control measures may be implemented where appropriate.
